## Solution {#sec-day1-sol .unnumbered}

### Part One:

Since `airquality` is always available, we do not need to import it explicitly like we will have to do usually. Let us assign this data to variable `df` and view it to help with understanding.

```{r, eval = F}
df <- airquality
class(df)

# View(df)
```

We can also see what `df` really is; it is a `data.frame`.

Now, there is an ABUNDANCE of ways to achieve the rest of the question. I will do a few:

-   **Brute Force:**

    This method uses for loops, considering every entry by going through each row and each column. We then check every entry to see if it is `NA`. If the entry is `NA` then take note of the row; otherwise check the next entry. Note: there is a `break` in the for loop because we do not want to double count a row if it has multiple columns which have `NA` entries for a specific row.

```{r}
# How do I know if something is na: is.na()
df <-  airquality

naRows <- c()
for (r in 1:nrow(df)){
  for (col in 1:ncol(df)){
    if (is.na(df[r, col])){
      naRows <- c(naRows, r)
      break
    }
  }
}
print(naRows)

```

-   **Progressive Elegance:**

```{r}

# which returns the indices which meet a certain condition
a <- c()
for (col in 1:ncol(df)){
  a <- c(a, which(is.na(df[, col])))
}
a <- sort(unique(a))

# complete.cases provides a boolean for whether the whole observation is NA free
b <- which(!complete.cases(df))

# Are these all the same?
cond1 <- all(a == b)
cond2 <- all(a == naRows)
cond3 <- all(b == naRows)
print("Are they the same thing?")
print(cond1 && cond2 && cond3)

```

### Part Two:

The unproblematic calculations are:

```{r}
min(df$Temp); max(df$Temp); mean(df$Temp); sd(df$Temp)
```

The *Ozone* data is a bit problematic because it contains `NA`'s which means that some of these calculations cannot be completed. We need to remove these `NA`'s but we could do this in two ways:

1.  Remove all incomplete rows.
2.  Only remove the `NA` for the ozone.

At this point, when we are just getting exposure to `R`, so there is no wrong way.

```{r}
# 1.
cleanDf <- df[-naRows, ]
min(cleanDf$Ozone); max(cleanDf$Ozone); mean(cleanDf$Ozone); sd(cleanDf$Ozone)

# 2.
cleanOzone <- na.omit(df$Ozone)
min(cleanOzone); max(cleanOzone); mean(cleanOzone); sd(cleanOzone)

# or even better:
min(df$Ozone, na.rm = T); max(df$Ozone, na.rm = T); mean(df$Ozone, na.rm = T); sd(df$Ozone, na.rm = T)
```

### Part Three:

Here we assign the response and covariate, and create the design matrix $X$. A design matrix may or may not be a familiar term to you. If not, a design matrix is the covariate matrix with a column of 1s. This is because it is required for the intercept term $\beta_0$.

```{r}
# Let us remove all the variables and clean the environment

rm(list = ls())

df <- cars
Y <- df$dist
speed <- df$speed
X <- cbind(1, speed)

b <- solve(t(X) %*% X) %*% t(X) %*% Y
print(b)

```

### Part Four:

```{r}
model <- lm(dist ~ speed, data = cars)
summary(model)

model$coefficients
```
