# Tidy Data and Reshaping {#sec-day4-tidydata}

In our introduction we have made reference to *Tidy Data* without really defining it. In this section, we will introduce the notion of tidy data, a tidyverse approach to dataframes, and reshaping dataframes between the long and wide representations.

## Base R Dataframes

Working with large dataframes can be challenging if you inadvertently print all rows to the console. Consider the following example:

```{r}
example_df <- data.frame(
  id = rep(paste0("id_", 1:500), each = 20),
  timepoint = rep(1:20, 500),
  disease_status = rep(
    sample(c("healthy", "diseased"), 20, replace = TRUE),
    each = 500
  ),
  blood_pressure = c(
    runif(20, 120, 140) |> round(),
    sample(c(NA, runif(19, 120, 140)), 9980, replace = TRUE)
  )
)
# Introduce a typo for demonstration
example_df[50, "disease_status"] <- "Healthy"
```

This dataframe contains 10,000 rows:

```{r}
nrow(example_df)
```

Printing the entire dataframe (e.g., simply typing `example_df`) would display all rows, which is both inconvenient and time-consuming. Instead, we typically use the `head()` function to preview just the first few rows:

```{r}
head(example_df)
```

However, constantly having to call `head()` is not ideal. A better solution is to convert the dataframe into a tibble.

## Tidyverse DataFrames

Since tidyverse aims to provide a consistent interface for the entire data workflow, it is only natural that an alternative to the base R dataframe is available. The alternative, *tibble,* can be thought of a modern dataframe which is very similar to the base version, with slight tweaks. A tibble dataframe is:

-   Lazy: It does not do any implicit conversions like converting strings to factors unless you tell it to

-   Strict: It will not allow you to anything ambiguous.

Whilst saying that a dataframe is lazy and strict does not sound like a positive, this prevents unexpected behaviour which you can easily sink hours into debugging.

Another benefit of tibble is an improved visualisation interface. When you print a tibble, it only shows the first 10 rows and includes data type information for each column. This is particularly useful because it lets you quickly verify, for example, whether `timepoint` is numeric or character; information that can be obscured in the default dataframe printout. This display can help you catch issues like typos or unexpected data types early on.

### Using `tibble`

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

Convert your dataframe to a tibble using `as_tibble()`:

```{r}
# Check the class before conversion
class(example_df)

# Convert to tibble
example_df <- as_tibble(example_df)

# Check the class after conversion
class(example_df)
```

Now, simply typing `example_df` will display a neat summary of your data:

```{r}
example_df
```

This tidy display helps you catch issues like typos or unexpected data types early on. It can be beneficial to load both `tibble` and `ggplot2` at the top of your scripts.

You can also create new dataframes directly as tibbles:

```{r}
new_tbl <- tibble(x = 1:5, y = rnorm(5))
new_tbl
```

Note that for a `tibble` object (i.e. a dataframe that has class `tbl_df`) to display as a tibble, the `tibble` package needs to be attached. So always add `library(tibble)` to the top of your script.

A final, very nice feature of `tibbles` is that selecting one column using `df[, "column"]` will return a tibble, rather than a vector. This is typically what we would expect:

```{r}
example_df[, "id"]
```


#### Exploring Data Columns with `view_cols`

Whilst `tibble`s provide a more informative display, they still only show the first few rows of each column.

Another useful tool is the `view_cols` function from the `UtilsDataRSV` package. This function displays unique entries for each column—always showing any missing values (`NA`s)—so you can quickly identify anomalies such as typos or unexpected values.

To install the `UtilsDataRSV` package, use the following code:

```{r}
#| eval: false
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("SATVILab/UtilsDataRSV")
```

Once installed, you can apply `view_cols` to your tibble to inspect the unique values in each column:

```{r}
UtilsDataRSV::view_cols(example_df)
```

`view_cols` is particularly helpful when you cannot easily inspect all the rows or columns of a dataframe. For example, with 10,000 rows, it’s impractical to scroll through the entire dataset to:

-   **Identify Typos:** In our example, the first ten rows may not reveal that `disease_status` contains three unique entries (e.g., `"healthy"`, `"diseased"`, and the typo `"Healthy"`).
-   **Detect Missing Data:** It's easy to overlook missing values (`NA`s).
-   **Verify Expected Values:** For instance, the `id` column should have 500 unique entries, and `view_cols` can help you confirm th

For a possibly more polished alternative, consider exploring the `skimr` package, which (I think) offers similar functionality.

You can also display the dataframe on its side, using `dplyr::glimpse()`:

```{r}
dplyr::glimpse(example_df)
```

## What is Tidy Data?

Tidy Data is defined a three strict rules. These rules define the structure of data, rather than allowing one's stylistic views to define how data is structured. A dataset is *tidy* if:

1.  Each **variable** has its own column
2.  Each **observation** has its own row
3.  Each **value** lives in a single cell

Consistent data structure simplifies analysis, leverages vectorized operations, and makes it easier to use tidyverse functions.

## Reshaping

Consider the following data. Is it tidy?

```{r}
# A classic "Wide" dataset
scores_wide <- tibble(
  Student = c("Alice", "Bob", "Charlie"),
  Math    = c(88, 72, 95),
  English = c(90, 85, 80),
  History = c(75, 88, 92)
)

print(scores_wide)
```

Recording data like this is very common in Excel and is a format we are familiar with. But what are the variables?

Converting the data to *tidy data* often improves how we can analyse (*filter, group, summarise, plot*) the data. We call this data *long* data, and the above Excel style data *wide* data. We can convert to long data with `pivot_longer()` from the `tidyr` package.

```{r}
scores_long <- scores_wide %>%
  pivot_longer(
    cols = c(Math, English, History), # The columns we want to bundle up
    names_to = "Subject",             # What should we call the new key column?
    values_to = "Score"               # What should we call the new value column?
  )

print(scores_long)
```

In this long format, every row is a single test result or a single observation. We can easily summarise the data now; for example:

```{r}
scores_long %>% 
  group_by(Student) %>% 
  summarise(Mean =  mean(Score))

scores_long %>% 
  group_by(Subject) %>% 
  summarise(Mean =  mean(Score))
```

If we want to return to the wide version of the data, for example if we have finished the analysis and want to create a table **OR** we want to cover calculations based on the **difference** between two subjects (this will be easier when the columns are still side-by-side), then we use `pivot_wider()`.

```{r}
# Let's turn our long data back into a human-readable table
report_table <- scores_long %>%
  pivot_wider(
    names_from = Subject,  # Which column holds the new column headers?
    values_from = Score    # Which column holds the values to fill the cells?
  )

print(report_table)
```

## Joining dataframes

The `dplyr` package also provides functions for joining dataframes, such as `left_join()` and `inner_join()`.

The full-length notes are in the [R4DS chapter on `joins`](https://r4ds.hadley.nz/joins).

### Key Joining Concepts in dplyr

Data analysis typically involves combining multiple data frames. Joins let you connect tables using shared keys, which can be primary keys (unique identifiers in one table) and foreign keys (variables that reference primary keys in another).

### Summary

Again, here is a concise summary of the R4DS content on the `joins` section. More examples are available in the help files for each function (e.g. `?left_join`).

To run the examples below, attach the `dplyr` package and create and load the following datasets:

```{r}
library(dplyr)
data("flights", package = "nycflights13")
flights2 <- flights |> 
  select(year, time_hour, origin, dest, tailnum, carrier)
data("airlines", package = "nycflights13")
data("planes", package = "nycflights13")
data("airports", package = "nycflights13")
df1 <- tibble(key = c(1, 2, 2), val_x = c("x1", "x2", "x3"))
df2 <- tibble(key = c(1, 2, 2), val_y = c("y1", "y2", "y3"))
```

Here is what these datasets look like:

```{r}
flights2
```

```{r}
airlines
```

```{r}
planes
```

```{r}
df1
```

```{r}
df2
```

#### Mutating Joins

Mutating joins add new columns from one data frame to another based on matching key values. They share a common interface:

-   **left_join()**: Keeps all rows from the left table and adds matching columns from the right table.

    ```{r}
    # Add full airline names to flights2 data
    flights2 |> left_join(airlines)
    ```

-   **inner_join()**: Keeps only rows with matching keys in both tables.

    ```{r}
    # Only keep rows where both x and y have a matching key
    df1 |> inner_join(df2)
    ```

    -   In the example above, `dplyr` emits a warning as two rows in `df1` (those with `key` value `2`) have two matches in `df2` (those with `key` value `2`). Often, this is an error (your data isn't what you expect it to be), but that is the example they showed.

-   **Specifying Keys:**

    By default, join functions match on columns with the same name. Use `join_by()` to specify different keys:

    ```{r}
    flights2 |> left_join(planes, join_by(tailnum))
    ```

#### Filtering Joins

Filtering joins select rows from one table based solely on whether they have a match in another table.

-   **semi_join()**: Keeps rows in `x` that have at least one match in `y` (does not add columns from `y`).

    ```{r}
    # Keep only origin airports that appear in flights2
    airports |> semi_join(flights2, join_by(faa == origin))
    ```

-   **anti_join()**: Keeps rows in `x` that have no match in `y`.

    ```{r}
    # Find tail numbers in flights2 that are missing from planes
    flights2 |> anti_join(planes, join_by(tailnum)) |> distinct(tailnum)
    ```

These joining functions provide a powerful and flexible way to integrate data from different sources, ensuring that your analyses are both comprehensive and accurate.

## TL;DR

1.  **Tibbles** are strict, lazy data frames that print nicely and prevent silent errors.

2.  **Wide Data** is good for humans (spreadsheets). **Long Data** is good for computers (*R*).

3.  Use `pivot_longer()` to bundle multiple columns into two key columns (e.g., `Subject` and `Score`).

4.  Use `pivot_wider()` to spread two columns back out into a table.
